---
title: "**Assignment 1  \n  K-NN regression  \n**"
author: "Andreu Meca, Geraldo Gariza, Victor Duque and Jan Leyva  \n   Universitat Politècnica de Catalunya" 
output:
  pdf_document: default
  html_document: default

date: "February 15, 2021"
fontsize: 12pt
linestretch: 1.5
toc-depth: 1
header-includes:
- \newcommand{\bcenter}{\begin{center}}
- \newcommand{\ecenter}{\end{center}}


---


```{r setup, include=TRUE, echo=F}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```
\pagenumbering{gobble}

\newpage


\pagenumbering{arabic}


\bcenter
# K-NN regression
\ecenter

The k nearest-neighbor estimator of $m(t)$ = E(Y|X = $t$) is defined as

$$\hat{m}(t)=\frac{1}{k}\sum_{i\in N_k(t)} y_i$$

where N~k~($t$) is the neighborhood of $t$ defined by the $k$ closest points $x$~$i$~ in the training sample.

### Boston housing data from library MASS:

```{r echo=TRUE}
library(MASS)
help(Boston)
data(Boston)
x <- Boston$lstat
y <- Boston$medv
plot(x,y, pch = 10,col = "#00BCF4")
```

### 1. Write a function for computing the $k$-nn estimator of $m$($t$) for a given value of t $\in$ $R$.

```{r echo=TRUE}
k<-5
t<-10
# dist matrix of X's
d_x   <- as.matrix(dist(rbind(t,as.matrix(x))))[1,-1]
d_x_k <- sort(d_x, partial=k)[k]
N_x_k <- unname( which(d_x <= d_x_k) )
(hat_y = sum(y[N_x_k])/k)
```

\newpage
#### Function $k$-nn for regression code
```{r echo=TRUE}
knn.reg <- function(t, x, y,  k=5){
  d_x   <- as.matrix(dist(rbind(t,as.matrix(x))))[1,-1]
  d_x_k <- sort(d_x, partial=k)[k]
  N_x_k <- unname( which(d_x <= d_x_k) )
  return(hat_y = sum(y[N_x_k])/k)
}
```


### 2. Then, define t as a sequence from 1 to 40: t <- 1:40.
```{r}
t <- 1:40
```

### 3. Estimate $m$(t[i]) for $i$ = 1,...,40 using $k$ = 50.
```{r echo=TRUE}
nt <- length(t)
hat_p <- matrix(0,1,ncol=nt)

for (i in 1:nt){
    hat_p[,i] <- knn.reg(t[i], x=x, y=y, k = 50)
}
```

\newpage
### 4. Plot y against x. Then represent the estimated regression function.
```{r  echo=TRUE}
plot(x, y, pch = 10,xlab = "lstat", ylab = "medv", col = "#00BCF4")
points(t, hat_p, col="#F8766D" , lwd=3 , pch=19 , type= "l")
points(t, hat_p, col="#F8766D" , lwd=3 , pch=19 , lty = "dashed", type= "b")
title("K-NN regression [K = 50]")
```
\newpage

### 5. Repeat the same exercise using different values of $k$.
```{r  echo=TRUE, figures-side, fig.show="hold", out.width="50%"}
for (n in 1:10) {
  a <- seq(from = 1, to = 55, by = 6)
  b <- seq(from = 6, to = 60, by = 6)
  k <- seq(a[n],b[n], by = 2)
  t <- 1:40
  hat_p.2 <- matrix(0,1, ncol = length(k), nrow = length(t))
  
  for (i in 1:length(t)) {
    for (j in 1:length(k)) {
      hat_p.2[i,j] <- knn.reg(t[i], x=x, y=y, k = k[j])
    }
  }
  # Plot all:
  par(mar = c(4, 4, 1.5, 1.5))
    for (i in 1:length(k)) {
      plot(x, y, pch = 10,xlab = "lstat", ylab = "medv", col = "#00BCF4")
      points(t, as.matrix(hat_p.2[,i]), col="#F8766D" , lwd=3 , pch=19 , type= "l")
      points(t, hat_p.2[,i], col="#F8766D" , lwd=3 , pch=19 , type= "b")
      title(paste("K =", k[i]))
    }
}
```


